Viral Sequence Sorting Tools Evaluation

### 2022-10-29 EnvironmentalData/DrinkingWaterMetagenomes/identifying_viral_contigs_environmental_dw.Rmd
Updating provirus rules to pull from all tools.

VIBRANT
    Quality == "High Quality Draft": +1
    Quality == "Medium Quality Draft": +1
    Quality == "Low Quality Draft" & provirus == TRUE: +0.5

Virsorter2
    Viral >= 50: +0.5
    Viral >= 0.95: +1
    Hallmark > 2: +1

Virsorter
    category ==  1,2,4,5: +1
    category == 3,6: +0.5

@@

Kaiju:
    Kaiju_viral = "cellular organisms": -1
    Kaiju_viral = "Viruses": +1

CheckV
    viral_genes == 0 and host_genes >= 1: keep_score = 0
    If 3*viral_genes <= host_genes: keep_score = 0
    If length > 50,000 and hallmark == 0: keep_score = 0
    If %unknown >= 75: +0.5

@@

```{r}
ggplot(num_viruses, aes(x=numtools, y=num_viruses)) +
  geom_boxplot(aes()) +
  geom_point(aes(color=assembly, fill=assembly)) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Number of Tools") +
  ylab("Num Viruses Predicted")
```

>>>>

VIBRANT
    Quality == "Complete Circular": +1
    Quality == "High Quality Draft": +1
    Quality == "Medium Quality Draft": +1
    Quality == "Low Quality Draft" & provirus == TRUE: +0.5

Virsorter2
    Viral >= 50: +0.5
    Viral >= 0.95: +0.5
    RNA >= 0.9: +1
    lavidaviridae >= 0.9: +1
    NCLDV >= 0.9: +1

Virsorter
    category ==  1,4: +1
    category == 2,5: +0.5

@@

Tuning - No Viral Signature:
    Kaiju_viral = "cellular organisms": -0.5
    If host_genes >50 and NOT checkv_provirus: -1 
    If viral_genes == 0 and host_genes >= 1: -1
    If 3*viral_genes <= host_genes and NOT checkv_provirus: -1
    If length > 50,000 and hallmark <=1: -1
    If length < 5000 and checkv completeness <= 75: -0.5

Tuning - Viral Signature:
    Kaiju_viral = "Viruses": +1
    If %unknown >= 75 and length < 50000: +0.5
    If %viral >= 50: +0.5
    Hallmark > 2: +1

@@

### 2022-10-29 TestingPipeline/Scripts/combining_tool_output.Rmd
Updating provirus rules to pull from all tools.

# Kaiju
viruses$Kaiju_Viral[is.na(viruses$Kaiju_Viral)] <- "unknown"
viruses$Kingdom[is.na(viruses$Kingdom)] <- "unknown"
```

>>>>

# Kaiju
viruses$Kaiju_Viral[is.na(viruses$Kaiju_Viral)] <- "unknown"
viruses$Kingdom[is.na(viruses$Kingdom)] <- "unknown"

#adding a combined provirus category
viruses$provirus <- FALSE
viruses$provirus[viruses$checkv_provirus=="Yes" | viruses$vibrant_prophage=="Yes" | viruses$category==4] <- TRUE
```

### 2022-10-29 TestingPipeline/Scripts/rule_tuning_2022.09.06.Rmd
Updating provirus rules to pull from all tools.

VIBRANT
    Quality == "High Quality Draft": +1
    Quality == "Medium Quality Draft": +1
    Quality == "Low Quality Draft" & provirus == TRUE: +0.5

Virsorter2
    Viral >= 50: +0.5
    Viral >= 0.95: +0.5

Virsorter
    category ==  1,2,4,5: +1
    category == 3,6: +0.5

DeepVirFinder:
    Score >= 0.7: +0.5
    Score >= 0.9: +0.5

Kaiju:
    Kaiju_viral = "cellular organisms": -1
    Kaiju_viral = "Viruses": +1

CheckV
    If %unknown >= 75: +0.5
    Hallmark > 2: +1
    viral_genes == 0 and host_genes >= 1: keep_score = 0
    If 3*viral_genes <= host_genes: keep_score = 0
    If length > 50,000 and hallmark == 0: keep_score = 0

@@

  if (include_vibrant) {
    keep_score[input_seqs$vibrant_quality=="complete circular"] <- keep_score[input_seqs$vibrant_quality=="complete circular"] + 1
    keep_score[input_seqs$vibrant_quality=="high quality draft"] <- keep_score[input_seqs$vibrant_quality=="high quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="medium quality draft"] <- keep_score[input_seqs$vibrant_quality=="medium quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$checkv_provirus=="Yes"] <- keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$checkv_provirus=="Yes"] + 0.5
  }

@@

  if (include_tuning_not_viral) {
    keep_score[input_seqs$Kaiju_Viral=="cellular organisms"] <- keep_score[input_seqs$Kaiju_Viral=="cellular organisms"] - 0.5
    keep_score[input_seqs$checkv_host_genes>50 & input_seqs$checkv_provirus=="No"] <- keep_score[input_seqs$checkv_host_genes>50 & input_seqs$checkv_provirus=="No"] - 1
    keep_score[input_seqs$checkv_viral_genes==0 & input_seqs$checkv_host_genes>=1] <- keep_score[input_seqs$checkv_viral_genes==0 & input_seqs$checkv_host_genes>=1] - 1
    keep_score[((input_seqs$checkv_viral_genes*3) <= input_seqs$checkv_host_genes) & input_seqs$checkv_provirus=="No"] <- keep_score[((input_seqs$checkv_viral_genes*3) <= input_seqs$checkv_host_genes) & input_seqs$checkv_provirus=="No"] - 1 # consider accounting for RNA viruses
    keep_score[input_seqs$checkv_length>500000 & input_seqs$hallmark<=1] <- keep_score[input_seqs$checkv_length>500000 & input_seqs$hallmark<=1] - 1
    keep_score[input_seqs$checkv_completeness<=75 & input_seqs$checkv_length<=5000] <- keep_score[input_seqs$checkv_completeness<=75 & input_seqs$checkv_length<=5000] - 0.5 # helped with protist contamination
  }

@@

```{r}
write_tsv(accuracy_scores, "20220927_accuracy_scores.tsv")
```

@@

accuracy:
```{r}
length(grep("true", viruses$confusion_matrix_high_recall))/nrow(viruses)
```

@@

>>>>

VIBRANT
    Quality == "Complete Circular": +1
    Quality == "High Quality Draft": +1
    Quality == "Medium Quality Draft": +1
    Quality == "Low Quality Draft" & provirus: +0.5

Virsorter2
    Viral >= 50: +0.5
    Viral >= 0.95: +0.5
    RNA >= 0.9: +1
    lavidaviridae >= 0.9: +1
    NCLDV >= 0.9: +1

Virsorter
    category ==  1,4: +1
    category == 2,5: +0.5

DeepVirFinder:
    Score >= 0.7: +0.5

Tuning - No Viral Signature:
    Kaiju_viral = "cellular organisms": -0.5
    If host_genes >50 and NOT provirus: -1 
    If viral_genes == 0 and host_genes >= 1: -1
    If 3*viral_genes <= host_genes and NOT provirus: -1
    If length > 50,000 and hallmark <=1: -1
    If length < 5000 and checkv completeness <= 75: -0.5

Tuning - Viral Signature:
    Kaiju_viral = "Viruses": +1
    If %unknown >= 75 and length < 50000: +0.5
    If %viral >= 50: +0.5
    Hallmark > 2: +1

@@

  if (include_vibrant) {
    keep_score[input_seqs$vibrant_quality=="complete circular"] <- keep_score[input_seqs$vibrant_quality=="complete circular"] + 1
    keep_score[input_seqs$vibrant_quality=="high quality draft"] <- keep_score[input_seqs$vibrant_quality=="high quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="medium quality draft"] <- keep_score[input_seqs$vibrant_quality=="medium quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$provirus] <- keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$provirus] + 0.5
  }

@@

  if (include_tuning_not_viral) {
    keep_score[input_seqs$Kaiju_Viral=="cellular organisms"] <- keep_score[input_seqs$Kaiju_Viral=="cellular organisms"] - 0.5
    keep_score[input_seqs$checkv_host_genes>50 & !input_seqs$provirus] <- keep_score[input_seqs$checkv_host_genes>50 & !input_seqs$provirus] - 1
    keep_score[input_seqs$checkv_viral_genes==0 & input_seqs$checkv_host_genes>=1] <- keep_score[input_seqs$checkv_viral_genes==0 & input_seqs$checkv_host_genes>=1] - 1
    keep_score[((input_seqs$checkv_viral_genes*3) <= input_seqs$checkv_host_genes) & !input_seqs$provirus] <- keep_score[((input_seqs$checkv_viral_genes*3) <= input_seqs$checkv_host_genes) & !input_seqs$provirus] - 1 # consider accounting for RNA viruses
    keep_score[input_seqs$checkv_length>500000 & input_seqs$hallmark<=1] <- keep_score[input_seqs$checkv_length>500000 & input_seqs$hallmark<=1] - 1
    keep_score[input_seqs$checkv_completeness<=75 & input_seqs$checkv_length<=5000] <- keep_score[input_seqs$checkv_completeness<=75 & input_seqs$checkv_length<=5000] - 0.5 # helped with protist contamination
  }

@@

```{r}
write_tsv(accuracy_scores, "20221029_accuracy_scores.tsv")
```

@@

accuracy:
```{r}
length(grep("true", viruses$confusion_matrix_high_recall))/nrow(viruses)
```
0.887

@@

0.960
